### 2022-02-18 pre/pre.tex
pres

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We finish with the description of the
    superposition calculus.
\end{abstract}

@@

Our work mostly concerns retrofitting first-order provers to support
higher-order concepts. But part of our work also concerns integrating
successful approaches for propositional logic in first-order
provers. We begin by defining all three logics, in the increasing
order of expressivity.

@@

@@

Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$.

@@

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
as $\sigmacl{l}$ and $\sigmacl{C}$.

@@

@@

>>>>

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by describing the three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We finish with the
    description of the superposition calculus. The text of this chapter is
    partially based on the preliminaries sections of the publications listed in
    Chapter \ref{ch:intro}.

\end{abstract}

@@

@@

Even though we do not formally define semantics of logic, we assume the natural
extensions of domain, valuation, interpretation and model (as defined by Fitting
\cite{mf-1996-fol}) from unsorted to many-sorted logic. Models we consider are
\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
of (un)satisfiability and (in)validity are assumed. We write
$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
(i.e., it satisfies it under every variable assignment), we simply write
$\ourmodel \models N.$  Abusing notation, we write $M \models N$ to denote that
$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$

@@

Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$ results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma t\right)$. We say that substitution is grounding
if it maps all variables to ground terms.

@@

Applying substitution to a literal is reduced to applying it to both sides of
the (dis)equa\-tion and it is extended pointwise to clauses. It is written in postfix notation
as $\sigmacl{l}$ and $\sigmacl{C}$. We say $\sigmacl{C}$ is a ground instance of $C$ is $\sigma$ is a grounding
substitution.

@@

Above described rules are used to generate new clauses from already derived
ones. Even though orderings and selection are used to reduce the search space,
it still grows rather fast. To identify unnecessary clauses from the search
space, superposition features \emph{redundancy criterion}
\cite[Sect.~4.2.2]{bg-01-resolution}. We say that a ground clause $C$ is
redundant in a ground set of clauses $N$ if it is entailed by a subset of
clauses in $N$ such that each clause in this subset is smaller than $C$. More
generally, a clause (ground or not) is redundant in set $N$ if every ground
instance of $C$ is redundant for some grounding of $N$. Removing redundant
clauses does not affect completeness of superposition.

Other than the four generating rules, superposition provers also use
simplification rules, which are justified using redundancy criterion. These
rules replace the premises with conclusions, and we denote them using double
bars. As an example of such rule, let us introduce rewriting (demodulation) of
negative literals (\infname{RN}):

\[
\namedinference{RN}{s \eq t \qquad u \noteq v \llor D}{(u[p \leftarrow \sigma(t)] \eq v \llor C \llor D)\sigma}
\]
where $\sigma = \mgu(u|_p, s)$, and $\sigma(s) \succ \sigma(t)$.

Simple rules such as removing duplicate literals, literals of the form $s \noteq
t$ or tautological clauses are clearly justified by the redundancy criterion.
Schulz gives an extensive list of such rules\cite{ss-02-brainiac}.

We say that a clause $C$ subsumes clause $D$ is there is a substitution $\sigma$
such that $\sigmacl{C} \subseteq D$. Subsumption is one of the most important
operations within a theorem prover. This rule does not adhere to redundancy
criterion, but can be justified using other mechanisms \cite{wtrb-20-sat-framework}.

@@

Superposition provers saturate the input problem with respect to the calculus's
inference rules using the \emph{given clause procedure}
\cite{mcw-1997-otter,adf-1995-discount}. It partitions the proof state into a
passive set $\mathcal{P}$ and an active set $\mathcal{A}$. All clauses start in
$\mathcal{P}$. At each iteration of the procedure's main loop, the prover
chooses a clause $C$ from $\mathcal{P}$, simplifies it, and moves it to
$\mathcal{A}$. Then all inferences between $C$ and active clauses are performed.
The resulting clauses are again simplified and put in $\mathcal{P}$.
The provers differ in which clauses are used for simplification: Otter-loop
\cite{mcw-1997-otter} provers use both active and passive clauses whereas
DISCOUNT-loop \cite{adf-1995-discount} provers use only active clauses.
The provers we discuss in this thesis are both DISCOUNT-loop provers.
